---
title: "edgeR Babesia RNAseq analysis"
output: pdf_document
---

```{r setup, include=FALSE}

      knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

      library(purrr)
      library(tidyverse)
      library(data.table)
      library(plotly)
      library(ggrepel)
      library(edgeR)
      library(magrittr)
      library(stringi)
      library(biomaRt)
      library(topGO)
      library(GO.db)
      library(reactome.db)
      library(kableExtra)
      library(RColorBrewer)
      library(matrixStats)
      library(cowplot)
      library(ggpubr)
      library(pheatmap)

      load("objects/analysis_data.Rdata")
      load("objects/degs.Rdata")
      load("objects/cluster_plots.Rdata")
      load("objects/pathways.Rdata")
      load("objects/clinicalfigures.Rdata")


      source("functions/volcano_plot.R")
      source("functions/pathways_plot.R")

```

### Figure 1: Clinical parameters

```{r format_data_function}

      format_clin_data <- function(df){
         
         df1 <- df %>% 
            pivot_longer(cols = 3:ncol(.), names_to = "day", values_to = "value" ) 
         
         
         df1$day <- gsub("[^0-9.-]", "", df1$day) %>% as.factor()
         df1$`Exp group`[is.na(df1$`Exp group`)] <- 0
         df2 <- df1 %>% mutate(inoculum = ifelse(`Exp group` == 0, "low", "high"))
         
         df2
         
         
      }   


```

```{r plot_data_function}

plot_clin_data <- function(df) {
      
      
      df %>% ggplot(aes(x = day, y = value, color = inoculum)) +
            geom_point(size = 4, alpha = 0.7) +
            stat_summary(aes(group = `Exp group`), fun = mean, geom = "line", size = 2) +
            xlim(c("-1","0", "1", "2", "3", "4", "5", "6", "7", "8")) +
            xlab("") +
            ylab("") +
            theme_classic() +
            theme(plot.title = element_text(hjust = 0.5, size = 30),
                  legend.position = "none",
                  axis.text.x = element_blank(),
                  axis.text.y = element_text(size = 25),
                  )
      
}


```

```{r vital_signs}

        # Temp
      
            p_temp <- df_temp %>% format_clin_data() %>% plot_clin_data() + ggtitle(expression("Temperature (" * degree * C * ")"))
      
        # Heart rate
      
            p_hr <- df_pulse %>% format_clin_data() %>% plot_clin_data() + ggtitle("Heart rate (BPM)") 
      
        # Blood pressure (systolic/diastolic)
            
            df_sys_bp1 <- df_sys_bp %>% format_clin_data() %>% mutate(type = "sys")
            df_dias_bp1 <- df_dias_bp %>% format_clin_data() %>% mutate(type = "dias")
            
            df_bp <- rbind(df_sys_bp1, df_dias_bp1)
            df_bp1 <- df_bp %>% 
                  mutate(group = paste0(inoculum, sep = "_", type))
            
            p_bp <- df_bp1 %>% ggplot(aes(x = day, y = value, color = inoculum)) +
                  geom_point(size = 4, alpha = 0.7) +
                  stat_summary(aes(group = group), fun = mean, geom = "line", size = 2) +

                  xlim(c("-1","2", "3", "4", "5", "6", "7", "8")) +
                  xlab("") +
                  ylab("") +
                  ggtitle("Blood pressure (sys/dias)") +
                  theme_classic() +
                  theme(plot.title = element_text(hjust = 0.5, size = 30),
                        legend.position = "none",
                        axis.text.x = element_blank(),
                        axis.text.y = element_text(size = 25),
                  )
            

        # Respiratory rate
            
            p_resp_rate <- df_resp_rate %>% format_clin_data %>% plot_clin_data() + ggtitle("Respiratory rate (BPM)")
      
        # Oxygen saturation
            
            p_oxy_sat <- df_oxy_sat %>% format_clin_data() %>% plot_clin_data() + ggtitle(expression(O[2] * " Saturation (%)")) 
      
        # Habitus
      
            p_habitus <- df_habitus %>% format_clin_data() %>% plot_clin_data() + ggtitle("Habitus (1-4)") 
            
        # Appetite
            
            p_appetite <- df_appetite %>% format_clin_data() %>% plot_clin_data() + ggtitle("Appetite (1-4)")


```

```{r hematology}

         # Parasitemia
            
            p_par <- df_par %>% format_clin_data() %>% plot_clin_data()  + ggtitle("Parasitemia (%)")

         # Hemoglobin
            
            p_Hb <- df_Hb %>% format_clin_data() %>% plot_clin_data() + ggtitle("Hemoglobin (g/L)")
      
         # Reticulocyte
            
            colnames(df_retic) <- c("Dog number", "Exp group", "-1", "0", "1", "2", "3", "4", "5", "6", "7", "8")
            df_retic2 <- df_retic %>% mutate(inoculum = substr(`Exp group`, start = 1, stop = nchar(`Exp group`) - 5))
            
            df_retic3 <- df_retic2 %>% pivot_longer(cols = 3:12, names_to = "day", values_to = "value" ) 
            
            p_retic <- df_retic3 %>% mutate(value = as.numeric(value)) %>% plot_clin_data() + ggtitle(expression(Reticulocytes ~ (10^9 /L)))
      
         # White blood cell
            
            p_white_cell <- df_white_cell %>% format_clin_data() %>% plot_clin_data() + ggtitle(expression(White ~ blood ~ cells ~ (10^9 /L)))
      
         # Band forms
            
            p_band_cell <- df_band_cell %>% format_clin_data() %>% plot_clin_data() + ggtitle(expression(Band ~ cells ~ (10^9 /L))) 
      
         # Platelets
            
            p_platelet <- df_platelet %>% format_clin_data() %>% plot_clin_data() + ggtitle(expression(Platelets ~ (10^9 /L)))
            
        # Neutrophil count
            
            p_neutrophil <- df_neutrophil %>% format_clin_data() %>% plot_clin_data() + ggtitle(expression(Neutrophils ~ (10^9 /L)))



```

```{r biochemistry}

         # plasma hemoglobin
            
            p_cfHb <- df_cfHb %>% format_clin_data() %>% plot_clin_data() + ggtitle("Plasma hemoglobin (g/L)") 
      
         # serum urea

            colnames(df_urea) <- c("Exp group", "-1", "0", "1", "2", "#", "3", "4", "5", "6", "7", "8")
            df_urea1 <- df_urea %>% dplyr::select(-c("#"))
            df_urea2 <- df_urea1 %>% mutate(inoculum = substr(`Exp group`, start = 1, stop = nchar(`Exp group`) - 5))
            
            df_urea3 <- df_urea2 %>% pivot_longer(cols = 2:9, names_to = "day", values_to = "value" ) 
            
            p_urea <- df_urea3 %>% mutate(value = as.numeric(value)) %>% plot_clin_data() + ggtitle("Serum urea (mmol/L)") 

      
         # serum creatinine
            
            p_creatinine <- df_creatinine %>% format_clin_data() %>% plot_clin_data() + ggtitle("Serum creatinine (µmol/L)")
      
         # blood HCO3
            
            df_hco3_1 <- df_hco3 %>% .[,1:10] %>% pivot_longer(cols = 3:ncol(.), names_to = "day", values_to = "value" )
            
            df_hco3_1$day <- gsub("HCO3", "", df_hco3_1$day)
            df_hco3_1$day <- gsub("[^0-9.-]", "", df_hco3_1$day) %>% as.factor()
            df_hco3_1$`Exp group`[is.na(df_hco3_1$`Exp group`)] <- 0
            df_hco3_2 <- df_hco3_1 %>% mutate(inoculum = ifelse(`Exp group` == 0, "low", "high"))
            
            p_hco3 <- df_hco3_2 %>% plot_clin_data() + ggtitle(expression(HCO[3] ~ (mmol/L)))
      
         # blood gluose
            
            p_glucose <- df_glucose %>% format_clin_data() %>% plot_clin_data() + ggtitle("Blood glucose (mmol/L)")
      
         # ALT/AST
            
            p_alt <- df_alt %>% format_clin_data() %>% plot_clin_data() + ggtitle("ALT (U/L)") 
            
          # CRP
            
            p_crp <- df_crp %>% format_clin_data() %>% plot_clin_data() + ggtitle("C-reactive protein (mg/L)")



```

```{r acid_base_endocrine}

         # pH
            
            p_pH <- df_pH %>% format_clin_data() %>% plot_clin_data() + ggtitle("pH")
      
         # pCO2
            
            df_pco2_1 <- df_pco2 %>% .[,1:10] %>% pivot_longer(cols = 3:ncol(.), names_to = "day", values_to = "value" )
            
            df_pco2_1$day <- gsub("CO2", "", df_pco2_1$day)
            df_pco2_1$day <- gsub("[^0-9.-]", "", df_pco2_1$day) %>% as.factor()
            df_pco2_1$`Exp group`[is.na(df_pco2_1$`Exp group`)] <- 0
            df_pco2_2 <- df_pco2_1 %>% mutate(inoculum = ifelse(`Exp group` == 0, "low", "high"))
            
            p_pco2 <- df_pco2_2 %>% plot_clin_data() + ggtitle(expression(pCO[2] ~ (mmHg))) 
      
         # pO2
            
            df_po2_1 <- df_po2 %>% .[,1:10] %>% pivot_longer(cols = 3:ncol(.), names_to = "day", values_to = "value" )
            
            df_po2_1$day <- gsub("O2", "", df_po2_1$day)
            df_po2_1$day <- gsub("[^0-9.-]", "", df_po2_1$day) %>% as.factor()
            df_po2_1$`Exp group`[is.na(df_po2_1$`Exp group`)] <- 0
            df_po2_2 <- df_po2_1 %>% mutate(inoculum = ifelse(`Exp group` == 0, "low", "high"))
            
            p_po2 <- df_po2_2 %>% plot_clin_data() + ggtitle(expression(pO[2] ~ (mmHg)))
      
         # lactate
            
            p_lactate <- df_lactate %>% format_clin_data() %>% plot_clin_data() + ggtitle("Lactate (mmol/L)") 
      
         # cortisol
            
            colnames(df_cortisol) <- c("Dog number", "Exp group", "-1", "0", "1", "2", "3", "4", "5", "6", "7", "8")
            df_cortisol1 <- df_cortisol %>% mutate(inoculum = substr(`Exp group`, start = 1, stop = nchar(`Exp group`) - 5))
            
            df_cortisol2 <- df_cortisol1 %>% pivot_longer(cols = 3:12, names_to = "day", values_to = "value" ) 
            
            p_cortisol <- df_cortisol2 %>% mutate(value = as.numeric(value)) %>% plot_clin_data() + ggtitle("Cortisol (nmol/L)")

         # T4
            
            colnames(df_t4) <- c("Dog number", "Exp group", "-1", "0", "1", "3", "4", "5", "6", "8")
            df_t4_1 <- df_t4 %>% mutate(inoculum = substr(`Exp group`, start = 1, stop = nchar(`Exp group`) - 5))
            
            df_t4_2 <- df_t4_1 %>% pivot_longer(cols = 3:10, names_to = "day", values_to = "value" ) 
            
            p_t4 <- df_t4_2 %>% mutate(value = as.numeric(value)) %>% plot_clin_data() + ggtitle("Total T4 (nmol/L)") 



```

```{r legend, fig.height = 1, fig.align="right", out.width = "50%"}

      p_filler <- df_temp %>%
            format_clin_data() %>% 
            ggplot(aes(x = day, y = value, color = inoculum)) +
            geom_point(size = 4, alpha = 0.7) +
            stat_summary(aes(group = `Exp group`), fun = mean, geom = "line", size = 2) +
            theme_classic()


      p_legend <- get_legend(p_filler)

      as_ggplot(p_legend)

```

```{r titles}

df_titles <- data.frame("Vital signs", "Hematology", "Acid/base and endocrine", "Biochemistry")

kable(df_titles, "latex", booktabs = TRUE, col.names = NULL, align = rep("c", 4)) %>% 
    column_spec(1:4, width = "3.85cm", border_left = TRUE, border_right = TRUE) %>% 
   kable_styling(font_size = 9)


```

```{r row1, fig.show = "hold", out.width="25%"}

      p_temp
      
      p_par
      
      p_pH
      
      p_cfHb
        
```

```{r row2, fig.show = "hold", out.width="25%"}

      p_hr

      p_Hb
      
      p_hco3

      p_urea
      
```

```{r row3, fig.show = "hold", out.width="25%"}

      p_bp

      p_retic
      
      p_pco2
      
      p_creatinine

```

```{r row4, fig.show = "hold", out.width="25%"}

      p_resp_rate

      p_white_cell
      
      p_po2
      
      p_glucose

```

```{r row5, fig.show = "hold", out.width="25%"}

      p_oxy_sat

      p_neutrophil
      
      p_lactate
      
      p_alt

```

```{r row6, fig.show = "hold", out.width="25%"}

      p_habitus 

      p_band_cell
      
      p_cortisol
      
      p_crp + xlab("day") + theme(axis.title.x = element_text(size = 30), axis.text.x = element_text(size = 25)) 
     
        
```

```{r row7, fig.show = "hold", out.width="25%"}

      p_appetite + xlab("day") + theme(axis.title.x = element_text(size = 30), axis.text.x = element_text(size = 25))

      p_platelet + xlab("day") + theme(axis.title.x = element_text(size = 30), axis.text.x = element_text(size = 25))
      
      p_t4 + xlab("day") + theme(axis.title.x = element_text(size = 30), axis.text.x = element_text(size = 25))
      
```

**Figure 1: Clinical data recorded throughout experimental time course in both the low and high inoculum cohorts.** Habitus was scored 1 -4 with 1 being complete collapse and 4 being a normal bright and alert dog. BPM: beats per minute. Plots rendered using the R package ggplot.

### Figure 2: Cytokine concentrations

```{r fig2_plots}

      # GM-CSF

            colnames(df_gm_csf) <- c("Dog number", "Exp group", "-1", "0", "1", "2", "3", "4", "5", "6", "7", "8")
            
            p_gm_csf <- df_gm_csf %>% format_clin_data() %>%  mutate(value = as.numeric(value)) %>% 
              
                ggplot(aes(x = day, y = value, color = inoculum)) +
                  geom_point(size = 4, alpha = 0.7) +
                  stat_summary(aes(group = `Exp group`), fun = mean, geom = "line", size = 2) +
                  xlim(c("-1","0", "1", "2", "3", "4", "5", "6", "7", "8")) +
                  xlab("") +
                  ylab("") +
                  theme_classic() +
                  theme(plot.title = element_text(hjust = 0.5, size = 30),
                        legend.position = "none",
                        axis.text.x = element_blank(),
                        axis.text.y = element_text(size = 25),
                        ) +
                  ggtitle("GM-CSF (pg/mL)")

      # Keratinocyte chemotactic-like cytokine
            
            colnames(df_keratinocyte) <- c("Dog number", "Exp group", "-1", "0", "1", "2", "3", "4", "5", "6", "7", "8")
            
            p_keratinocyte <- df_keratinocyte %>% 
                  format_clin_data() %>%  
                  mutate(value = as.numeric(value)) %>% 
                  plot_clin_data() + ggtitle("KC-like cytokine (pg/mL)") 


      # IL-6 
            
            colnames(df_il6) <- c("Dog number", "Exp group", "-1", "0", "1", "2", "3", "4", "5", "6", "7", "8")
            
            p_il6 <- df_il6 %>% format_clin_data() %>%  mutate(value = as.numeric(value)) %>% 
              
              ggplot(aes(x = day, y = value, color = inoculum)) +
                  geom_point(size = 4, alpha = 0.7) +
                  stat_summary(aes(group = `Exp group`), fun = mean, geom = "line", size = 2) +
                  xlim(c("-1","0", "1", "2", "3", "4", "5", "6", "7", "8")) +
                  xlab("") +
                  ylab("") +
                  theme_classic() +
                  theme(plot.title = element_text(hjust = 0.5, size = 30),
                        legend.position = "none",
                        axis.text.x = element_blank(),
                        axis.text.y = element_text(size = 25),
                  ) +
                  ggtitle("IL-6 (pg/mL)") 
            

      # IL-8
            
            colnames(df_il8) <- c("Dog number", "Exp group", "-1", "0", "1", "2", "3", "4", "5", "6", "7", "8")
            
            p_il8 <- df_il8 %>% format_clin_data() %>%  mutate(value = as.numeric(value)) %>% 
              
              ggplot(aes(x = day, y = value, color = inoculum)) +
                  geom_point(size = 4, alpha = 0.7) +
                  stat_summary(aes(group = `Exp group`), fun = mean, geom = "line", size = 2) +
                  xlim(c("-1","0", "1", "2", "3", "4", "5", "6", "7", "8")) +
                  xlab("") +
                  ylab("") +
                  theme_classic() +
                  theme(plot.title = element_text(hjust = 0.5, size = 30),
                        legend.position = "none",
                        axis.text.x = element_blank(),
                        axis.text.y = element_text(size = 25),
                  ) +
                  ggtitle("IL-8 (pg/mL)")
            
      # IL-10 
            
            colnames(df_il10) <- c("Dog number", "Exp group", "-1", "0", "1", "2", "3", "4", "5", "6", "7", "8")
            
            p_il10 <- df_il10 %>% format_clin_data() %>%  mutate(value = as.numeric(value)) %>% 
              
              ggplot(aes(x = day, y = value, color = inoculum)) +
                  geom_point(size = 4, alpha = 0.7) +
                  stat_summary(aes(group = `Exp group`), fun = mean, geom = "line", size = 2) +
                  xlim(c("-1","0", "1", "2", "3", "4", "5", "6", "7", "8")) +
                  xlab("") +
                  ylab("") +
                  theme_classic() +
                  theme(plot.title = element_text(hjust = 0.5, size = 30),
                        legend.position = "none",
                        axis.text.x = element_blank(),
                        axis.text.y = element_text(size = 25),
                  ) +
                  ggtitle("IL-10 (pg/mL)") 
            
      # Macrophage chemotactic protein-1 
            
            colnames(df_macrophage) <- c("Dog number", "Exp group", "-1", "0", "1", "2", "3", "4", "4.1", "5", "6", "7", "8")
            
            p_macrophage <- df_macrophage %>% mutate("8" = as.character(df_macrophage$"8")) %>% 
                  format_clin_data() %>%  
                  mutate(value = as.numeric(value)) %>% 
              
                  ggplot(aes(x = day, y = value, color = inoculum)) +
                      geom_point(size = 4, alpha = 0.7) +
                      stat_summary(aes(group = `Exp group`), fun = mean, geom = "line", size = 2) +
                      xlim(c("-1","0", "1", "2", "3", "4", "5", "6", "7", "8")) +
                      xlab("") +
                      ylab("") +
                      theme_classic() +
                      theme(plot.title = element_text(hjust = 0.5, size = 30),
                            legend.position = "none",
                            axis.text.x = element_blank(),
                            axis.text.y = element_text(size = 25),
                      ) +
                      ggtitle("MCP-1 (pg/mL)") 
            

      # TNF-alpha
            
            colnames(df_tnfa) <- c("Dog number", "Exp group", "-1", "0", "1", "2", "3", "4", "5", "6", "7", "8")
            
            p_tnfa <- df_tnfa %>% format_clin_data() %>%  mutate(value = as.numeric(value)) %>% 
              
              ggplot(aes(x = day, y = value, color = inoculum)) +
                  geom_point(size = 4, alpha = 0.7) +
                  stat_summary(aes(group = `Exp group`), fun = mean, geom = "line", size = 2) +
                  xlim(c("-1","0", "1", "2", "3", "4", "5", "6", "7", "8")) +
                  xlab("") +
                  ylab("") +
                  theme_classic() +
                  theme(plot.title = element_text(hjust = 0.5, size = 30),
                        legend.position = "none",
                        axis.text.x = element_blank(),
                        axis.text.y = element_text(size = 25),
                  ) +
                  ggtitle(expression(paste(~TNF~alpha ~(pg/mL))))
            

 

```

```{r fig2, fig.show = "hold", out.width="25%"}

      p_gm_csf 

      p_keratinocyte 
      
      p_il6 
      
      p_il8 + xlab("day") + theme(axis.title.x = element_text(size = 30), axis.text.x = element_text(size = 25))
      
      p_il10 + xlab("day") + theme(axis.title.x = element_text(size = 30), axis.text.x = element_text(size = 25))
      
      p_macrophage + xlab("day") + theme(axis.title.x = element_text(size = 30), axis.text.x = element_text(size = 25))
      
      p_tnfa + xlab("day") + theme(axis.title.x = element_text(size = 30), axis.text.x = element_text(size = 25))
      
        
```

**Figure 2: Plasma cytokine concentrations recorded throughout experimental time course in both the low and high inoculum cohorts.** 

### Figure 3: Sources of variation among samples

#### A: Pairwise correlation across samples
```{r transform_counts}

      # Variance stabilizing transformation to visualize counts
      m_rawcounts_filtered_trans <- df_rawcounts_filtered[3:30] %>% as.matrix()
      rownames(m_rawcounts_filtered_trans) <- df_rawcounts_filtered$ensgene
      m_rawcounts_filtered_trans <- log(m_rawcounts_filtered_trans + 1) # natural log, add constant to deal with 0s
      
      # Relabel columns
      df_metadata1 <- df_metadata %>% mutate(day = paste0("D",day))
      
      df_metadata2 <- df_metadata1 %>% mutate(dose_spl = paste0(dose, 
                                                                sep = ifelse(dose == "low",
                                                                             "__",
                                                                             "_"), 
                                                                case_when(
                                                                   
                                                                   dog == 0 ~ "0",
                                                                   dog == 1 ~ "1",
                                                                   dog == 2 ~ "2", 
                                                                   dog == 3 ~ "1",
                                                                   dog == 4 ~ "2", 
                                                                   dog == 5 ~ "3"
                                                                   
                                                                )))
      
      df_metadata3 <- df_metadata2 %>% mutate(label = paste0(dose_spl, sep = "_", day))
                                                
      colnames(m_rawcounts_filtered_trans) <- df_metadata3$label

```

```{r fig3_heatmap, fig.height = 7, fig.width = 7.5}

      # Correlate samples based on transformed counts
      m_cor_samples <- cor(m_rawcounts_filtered_trans)
      
      # Plot
      pheatmap(m_cor_samples)

```

#### B: Principal Components Analysis 
```{r fig3_pca}

      # Calculate PCA data
      l_pca_data <- prcomp(m_rawcounts_filtered_trans)
      df_pca_data <- l_pca_data$rotation %>% data.frame() %>% rownames_to_column("label") %>% as_tibble()
      df_pca_meta_data <- df_metadata3 %>% left_join(df_pca_data, by = "label")
      
      # Plot
      df_pca_meta_data %>% ggplot(aes(x = PC1, y = PC2)) +
         
         geom_point(aes(color = dose), size = 3) +
         geom_text_repel(aes(label = substr(day, start = 2, stop = 2)), color = "black") +
         
         labs(color = "inoculum") +
         
         xlab(paste0("PC1: ", round(summary(l_pca_data)$importance[2,1]*100, 2), "% of variance")) +
         ylab(paste0("PC2: ", round(summary(l_pca_data)$importance[2,2]*100, 2), "% of variance")) +
         
         theme_bw()


```

### Figure 4: Pairwise day-by-day comparisons

```{r inocula, out.width = "100%"}

      df_inoculum <- data.frame("Low inoculum", "High inoculum")
      
      kable(df_inoculum, "latex", booktabs = TRUE, col.names = NULL, align = rep("c", 2)) %>% 
          column_spec(1:2, width = "8cm", border_left = TRUE, border_right = TRUE) %>% 
         kable_styling(font_size = 15)

```

```{r volc_d1, fig.show = "hold", fig.height = 6, out.width = "50%"}

      volcano_plot(l_pairwise_l[1])

      volcano_plot(l_pairwise_h[1])

```

```{r paths_d1, fig.show = "hold", fig.height = 12, out.width = "50%"}

      pathways_plot(l_gorpaths_l[1])

      pathways_plot(l_gorpaths_h[1])

```

```{r volc_d3, fig.show = "hold", fig.height = 6, out.width = "50%"}

      volcano_plot(l_pairwise_l[2])

      volcano_plot(l_pairwise_h[2])

```

```{r paths_d3, fig.show = "hold", fig.height = 12, out.width = "50%"}

      pathways_plot(l_gorpaths_l[2])

      pathways_plot(l_gorpaths_h[2])

```

```{r volc_d4, fig.show = "hold", fig.height = 6, out.width = "50%"}

      volcano_plot(l_pairwise_l[3])

      volcano_plot(l_pairwise_h[3])

```

```{r paths_d4, fig.show = "hold", fig.height = 12, out.width = "50%"}

      pathways_plot(l_gorpaths_l[3])

      pathways_plot(l_gorpaths_h[3])

```

```{r volc_d6, fig.show = "hold", fig.height = 6, out.width = "50%"}

      volcano_plot(l_pairwise_l[4])

      volcano_plot(l_pairwise_h[4])

```

```{r paths_d6, fig.show = "hold", fig.height = 12, out.width = "50%"}

      pathways_plot(l_gorpaths_l[4])

      pathways_plot(l_gorpaths_h[4]) + ylim(1,105)

```

```{r volc_d8, fig.show = "hold", fig.height = 6, out.width = "50%"}

      volcano_plot(l_pairwise_l[5])

      volcano_plot(l_pairwise_h[5])

```

```{r paths_d8, fig.show = "hold", fig.height = 12, out.width = "50%"}

      pathways_plot(l_gorpaths_l[5]) + ylim(1,105)

      pathways_plot(l_gorpaths_h[5]) 

```

# Clusters

```{r low_clusters}

      # Determine which observations belong to which cluster    
      my_clusters_l <- cutree(tree = hclust_l, k = 3) 
      
      df_corr_l$cluster <- my_clusters_l
      df_clust_l <- df_corr_l %>% dplyr::select(rowname, cluster)
      names(df_clust_l)[1] = "ensgene"
      
      # plot
      
         # calculate number of genes in each cluster
         df_n_l <- df_clust_l %>% dplyr::count(cluster)
         df_clust_l <- df_clust_l %>% left_join(df_n_l, by = "cluster")
         
         # add cluster to low genes
         df_avg_l2fc_l <- df_avg_l2fc[, c(1,grep("low_*", colnames(df_avg_l2fc)))] %>% left_join(df_clust_l, by = "ensgene")
         
         # pivot
         df_avg_l2fc_pivot_l <- df_avg_l2fc_l %>% pivot_longer(cols = 2:7, names_to = "day", values_to = "L2FC") %>% 
            mutate(day = substr(day, start = 5, stop = 5))
         
         df_clust_day_avgl2fc_l <- df_avg_l2fc_pivot_l %>%
            
            mutate(n = as.numeric(n)) %>% 
            
            group_by(cluster, n, day) %>%
            
            summarise(mean_L2FC = mean(L2FC),
                      stdev_L2FC = sd(L2FC)) %>% 
            
            mutate(error = qnorm(0.975)*stdev_L2FC/sqrt(n)) %>% 
            mutate(left = mean_L2FC - error, right = mean_L2FC + error) %>% 
            
            mutate(day = as.numeric(day)) %>% 
            mutate(cluster = as.factor(cluster)) %>% 
            
            mutate(cluster_n = paste0(cluster, sep = ", n = ", n))
         
         # plot all clusters
         df_clust_day_avgl2fc_l %>% ggplot() + 
               geom_line(aes(x = day, y = mean_L2FC, col = cluster), size = 2) + 
               geom_errorbar(aes(x = day, ymin = left, ymax = right), width = 0.2) +
               facet_wrap(~cluster_n) +
               theme_minimal() +
               ggtitle("Low inoculum clusters") +
               theme(legend.position = "none")
         
```

```{r degs_l}

   # create list of genes
   l_degs_l <- 1:max(df_all_degs_l$cluster) %>%
      purrr::map(~df_all_degs_l %>% filter(cluster == .x) %>% pull(geneID) %>% paste(collapse = ", "))
   
   names(l_degs_l) <- c(1:max(df_all_degs_l$cluster))
   
   # convert to df
   df_degs_in_clust_l <- bind_rows(l_degs_l) %>%
      pivot_longer(cols = 1:max(df_all_degs_l$cluster), names_to = "cluster", values_to = "geneID")
   
   # pivot
   df_degs_in_clust_l <- df_degs_in_clust_l %>% pivot_wider(names_from = cluster, values_from = geneID)
   
   # print df
   kable(df_degs_in_clust_l, "latex", booktabs = TRUE, align = rep("c", 3)) %>% 
      column_spec(1:3, width = "5.1cm", border_left = TRUE, border_right = TRUE) %>% 
      kable_styling(font_size = 8)


```

```{r high_clusters}

      # Determine which observations belong to which cluster    
      my_clusters_h <- cutree(tree = hclust_h, k = 3) 
      
      df_corr_h$cluster <- my_clusters_h
      df_clust_h <- df_corr_h %>% dplyr::select(rowname, cluster)
      names(df_clust_h)[1] = "ensgene"
      
      # plot
      
         # calculate number of genes in each cluster
         df_n_h <- df_clust_h %>% dplyr::count(cluster)
         df_clust_h <- df_clust_h %>% left_join(df_n_h, by = "cluster")
         
         # add cluster to low genes
         df_avg_l2fc_h <- df_avg_l2fc[, c(1,grep("high_*", colnames(df_avg_l2fc)))] %>% left_join(df_clust_h, by = "ensgene")
         
         # pivot
         df_avg_l2fc_pivot_h <- df_avg_l2fc_h %>% pivot_longer(cols = 2:7, names_to = "day", values_to = "L2FC") %>% 
            mutate(day = substr(day, start = 6, stop = 6))
         
         df_clust_day_avgl2fc_h <- df_avg_l2fc_pivot_h %>%
            
            mutate(n = as.numeric(n)) %>% 
            
            group_by(cluster, n, day) %>%
            
            summarise(mean_L2FC = mean(L2FC),
                      stdev_L2FC = sd(L2FC)) %>% 
            
            mutate(error = qnorm(0.975)*stdev_L2FC/sqrt(n)) %>% 
            mutate(left = mean_L2FC - error, right = mean_L2FC + error) %>% 
            
            mutate(day = as.numeric(day)) %>% 
            mutate(cluster = as.factor(cluster)) %>% 
            
            mutate(cluster_n = paste0(cluster, sep = ", n = ", n))
         
         
         # plot all clusters
         df_clust_day_avgl2fc_h %>% ggplot() + 
               geom_line(aes(x = day, y = mean_L2FC, col = cluster), size = 2) + 
               geom_errorbar(aes(x = day, ymin = left, ymax = right), width = 0.2) +
               facet_wrap(~cluster_n) +
               theme_minimal() +
               ggtitle("High inoculum clusters") +
               theme(legend.position = "none")
         

```

```{r degs_h}

   # create list of genes
   l_degs_h <- 1:max(df_all_degs_h$cluster) %>%
      purrr::map(~df_all_degs_h %>% filter(cluster == .x) %>% pull(geneID) %>% paste(collapse = ", "))
   
   names(l_degs_h) <- c(1:max(df_all_degs_h$cluster))
   
   # convert to df
   df_degs_in_clust_h <- bind_rows(l_degs_h) %>%
      pivot_longer(cols = 1:max(df_all_degs_h$cluster), names_to = "cluster", values_to = "geneID")
   
   # pivot
   df_degs_in_clust_h <- df_degs_in_clust_h %>% pivot_wider(names_from = cluster, values_from = geneID)
   
   # print df
   kable(df_degs_in_clust_h, "latex", booktabs = TRUE, align = rep("c", 3)) %>% 
      column_spec(1:3, width = "5.1cm", border_left = TRUE, border_right = TRUE) %>% 
      kable_styling(font_size = 8)


```

### Overlapping DEGs

```{r overlap_genes}

   kable(df_all_degs, "latex", booktabs = TRUE, align = rep("l", 4), linesep = "") %>% 
      column_spec(1, width = "3.5cm", border_left = TRUE, border_right = TRUE) %>% 
      column_spec(2:4, width = "1.7cm", border_left = TRUE, border_right = TRUE) %>% 
      kable_styling(font_size = 8)


```









